---
title: "Annotate Cell Types and Save Individual Cell-Type ACTIONet Objects"
output:
  html_notebook: default
  pdf_document: default
---
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/home/samsl/Pseudotemporal-GRNs")
```

```{r "Load Libraries", warning=FALSE, include=FALSE, results='hide'}
require(ACTIONet)
require(SingleCellExperiment)
require(Matrix)
require(scater)
require(scran)
```

# Create SingleCellExperiment Object
If using an mtx file and metadata, create an sce object. If you already have an SCE, Seurat, or AnnData object, skip to the relevant section.
```{r "Create Single Cell Experiment"}
# Matrix and Metadata files
matrix.file.name = "data/Processed_10x_AD_data/filtered_count_matrix.mtx"
colMetadata.file.name = "data/Processed_10x_AD_data/filtered_column_metadata.txt"
idMapping.file.name = "data/Processed_10x_AD_data/id_mapping.csv"
rowNames.file.name = "data/Processed_10x_AD_data/filtered_gene_row_names.txt"

# Read in counts and metadata
filtered.counts <- readMM(matrix.file.name)
filtered.colMetadata <- read.delim(colMetadata.file.name)
filtered.colMetadata$pre.cluster <- factor(filtered.colMetadata$pre.cluster)

# Map cells to their samples
idMapping <- read.table(idMapping.file.name, header = TRUE, sep = ",", stringsAsFactors = FALSE)
filtered.colMetadata$Subject <- lapply(filtered.colMetadata$projid, function(x) { idMapping$Subject[[match(x,idMapping$projid)]] })

# Assign row names (genes) and column names (cell tags)
rownames(filtered.counts) <- readLines(rowNames.file.name)
colnames(filtered.counts) <- filtered.colMetadata$TAG

# Create tsne array
tsneDims <- filtered.colMetadata[c("tsne1", "tsne2")]

# Create SCE
colData <- filtered.colMetadata[c("projid","Subject","pre.cluster","broad.cell.type","Subcluster")]
sce <- SingleCellExperiment(assays = list(counts =filtered.counts), colData = colData)
reducedDim(sce) <- tsneDims
reducedDimNames(sce) <- "tsne"

```

## Processing
Normalize to log counts and perform batch correction with Harmony.
```{r "Reduce and Batch Correct"}
# Reduce and Batch Correct
if (is.null(sce$batch)) {
  sce <- reduce.sce(sce)
} else {
  sce <- reduce.and.batch.correct.sce.Harmony(sce, reduced_dim = 35, harmony_iters = 30, reduce_iters = 5)
}
```

Add gene metadata from biomart, including  ensembl symbol and position on the chromosome.
```{r "Biomart Annotation"}
# Add Biomart Information to Genes
require(biomaRt)
mart=useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")

bm.attributes <- c("ensembl_gene_id", "start_position", "end_position", "chromosome_name")
sce <- getBMFeatureAnnos(sce, filters="gene.id",
                            attributes = bm.attributes,
                            dataset = "hsapiens_gene_ensembl")
```

### Load from processed SCE Object
```{r "Load from SCE"}
# SCE file name
sce.file.name <- "data/sce.rds"

sce <- readRDS(sce.file.name)
```

### Load from Seurat Object
```{r "Load from Seurat"}
# Seurat File Name
seurat.file.name <- ""

sce <- import.sce.from.Seurat(readRDS(file = seurat.file.name))
```

### Load from AnnData File
```{r "Load from AnnData"}
# AnnData File Name
anndata.file.name <- ""

sce <- import.sce.from.AnnData(file = anndata.file.name)
```

## Save
Save the SCE for future analyses.
```{r "Save SCE", result='hide'}
sce.file.name <- "data/sce.rds"
saveRDS(sce, file = sce.file.name)
```

# Annotate Cell Types
Compute ACTIONet and use it to annotate cells according to their cell type given a set of cell type markers.
```{r "Compute ACTIONet"}
ACTIONet.out <- run.ACTIONet(sce, k_max = 20, thread_no = 12)
```
Select the appropriate cell type markers from the set of curated markers.
```{r "Select Markers"}
data("curatedMarkers_human")
markers <- curatedMarkers_human$Brain$PFC$major.celltypes$marker.genes
```
Run cell type annotation.
```{r "Annotate Cell Types Using Markers"}
# Annotate cell types
ACTIONet.out <- annotate.cells.from.archetypes.using.markers(ACTIONet.out,
                                                             markers,
                                                             annotation.name = "Cell.Types")
```
Visualize the cell types.
```{r "Plot cell types"}
plot.ACTIONet(ACTIONet.out, labels = ACTIONet.out$annotations$Cell.Types$Labels)
```
Save the full annotated ACTIONet object for later analyses.
```{r "Save annotated"}
# Full ACTIONet File
full.ACTIONet.file <- "data/allCellsACTIONet.rds"

# Save annotated full ACTIONet
saveRDS(ACTIONet.out, file = full.ACTIONet.file)
```

### Load Annotated ACTIONet
```{r "Load Annotated ACTIONet"}
ACTIONet.out <- readRDS(ACTIONet.out, file = full.ACTIONet.file)
```

## Save SCE and ACTIONet For Each Cell Type
We want to re-run ACTIONet for each subset of cells to learn archetypes within a cell type
```{r "Create ACTIONet for each cell type"}
cell.type.file.dir = "data"

for (cell.type in names(markers)) {
  print(paste("Running ACTIONet for",cell.type,"cells..."))
  keep.cells <- colnames(sce)[which(ACTIONet.out$annotations$Cell.Types$Labels == cell.type)]
  if (length(keep.cells) == 0) {
    print(paste("No cells labeled as ",cell.type))
    next 
  }
  cell.type.sce = sce[,keep.cells]
  cell.type.ACTIONet = run.ACTIONet(cell.type.sce, k_max = 20, thread_no = 12)
  saveRDS(cell.type.sce, file = paste(cell.type.file.dir,"/",cell.type,"_sce.rds",sep=""))
  saveRDS(cell.type.ACTIONet, file = paste(cell.type.file.dir,"/",cell.type,"_ACTIONet.rds",sep=""))
}

```
Now we can perform the analyses in constructPseudotemporalOrdering.Rmd using each `[cell.type]_sce.rds` and `[cell.type]_ACTIONet.rds`.