---
title: "learnDBN.Rmd"
output:
  html_notebook: default
  pdf_document: default
---
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/home/samsl/Pseudotemporal-GRNs")
knitr::opts_chunk$set(echo = TRUE)
```

```{r "Requirements",warning=FALSE, include=FALSE, results='hide'}
require(bnstruct)
require(dplyr)
require(Rgraphviz)

read.gmt <- function(file) {
  if (!grepl("\\.gmt$", file)[1]) {
    stop("Pathway information must be a .gmt file")
  }
  geneSetDB = readLines(file)
  geneSetDB = strsplit(geneSetDB, "\t")
  name = unlist(geneSetDB[1])
  desc = strsplit(geneSetDB[2][[1]],"> ")[[1]][2]
  geneSetDB = unlist(geneSetDB[-(1:2)])
  geneSetDB = list("name"=name,"description"=desc,"genes"=geneSetDB)
  return(geneSetDB)
}
```

```{r}
# Select which data to load
cell.type = "Mic"
clusts <- readRDS(paste0("processed-data/",cell.type,"/clusters.rds"))
df <- as.data.frame(readRDS(paste0("processed-data/",cell.type,"/top_gene_counts.rds")))
```

```{r}
# Load gene lists for network generation
marker.root <- "processed-data/markers/"
kegg <- read.gmt(paste0(marker.root,"KEGG_AD.gmt"))
adLib <- list("name"="library","description"="Library in AD SCE exeriment","genes"=colnames(df))

blalock_incpt_dn <- read.gmt(paste0(marker.root,"BLALOCK_INCPT_AD_DN.gmt"))
blalock_incpt_up <- read.gmt(paste0(marker.root,"BLALOCK_INCPT_AD_UP.gmt"))
blalock_dn <- read.gmt(paste0(marker.root,"BLALOCK_AD_DN.gmt"))
blalock_up <- read.gmt(paste0(marker.root,"BLALOCK_AD_UP.gmt"))

gene.sets <- list("kegg"=kegg$genes,"incpt_dn"=blalock_incpt_dn$genes,"incpt_up"=blalock_incpt_up$genes,"dn"=blalock_dn$genes,"up"=blalock_up$genes)

# Select subset of genes
gene.df <- df[,intersect(adLib$genes,kegg$genes)]

# Write output for GRNVBEM
pseudotime.sorted <- t(gene.df)[,order(clusts$DiseaseScore)]
write.csv(pseudotime.sorted, file=paste0("processed-data/",cell.type,"/GRNVBEM.csv"))
```

```{r}
# Divide cells into layers based on clusters
clusts$ScoreClusts = as.numeric(clusts$DiseaseRange)
K.clusters = length(unique(clusts$ScoreClusts))

layers = c()
for (i in seq(K.clusters)) {
  layers[[i]] <- as.data.frame(gene.df[clusts$ScoreClusts == i,])
  colnames(layers[[i]]) <- paste0(colnames(layers[[i]]),"_t",i)
}
```

```{r}
# Build bnstruct dataset
# Rows = observations = cells
# Columns = variables = genes
gene.names <- unlist(list(cbind(unlist(lapply(layers,colnames)))))

n <- sum(unlist(lapply(layers,function(x){dim(x)[1]})))

data <- NULL
for (i in seq(K.clusters))
{
  layer.n <- dim(layers[[i]])[1]
  layer.genes <- colnames(layers[[i]])
  padding <- t(replicate(n - layer.n, rep(NA,length(layer.genes))))
  colnames(padding) <- layer.genes
  padded.layer = rbind(layers[[i]], padding)
  if (is.null(data)){
      data <- padded.layer
    }
    else{
      data <- cbind(data,padded.layer)
    }
}
```

```{r}
BN.data <- BNDataset(data = data,
                     discreteness = rep('c', length(gene.names)),
                     variables = gene.names,
                     node.sizes = apply(data, 2, max, na.rm = TRUE)
                     )

BN.data <- impute(BN.data)
saveRDS(BN.data, paste0("processed-data/",cell.type,"/BNdata.rds"))


```

```{r}
BN.data <- readRDS(paste0("processed-data/",cell.type,"/BNdata.rds"))
net.layers <- apply(as.matrix(gene.names), MARGIN=1, FUN=function (x) {as.numeric(strsplit(x,"_t")[[1]][2])})

# Learn Network Structure
DBN <- learn.network(BN.data,
                     seed = 12345,
                     algo = "mmpc",
                     layering = net.layers)
saveRDS(DBN, paste0("processed-data/",cell.type,"/BN.rds"))
```

# output edges of graph
groups <- list()
for (i in seq(K.clusters)) {
  groups[[i]] <- names(network$nodes)[grep(paste0(".*_t",i),names(network$nodes))]
}
graphviz.plot(network,groups=groups)

# Adjacency Matrix
adjacency <- amat(network)

# SIF
write.sif <- function(file,network,df) {
  arcs = arc.strength(network, bndf)
  arcs$relation = rep.int("->",dim(arcs)[1])
  arcs$parentTime = apply(as.matrix(arcs$from), MARGIN=1, FUN=function (x) {as.numeric(strsplit(x,"_t")[[1]][2])})
  arcs = arcs[,c("from","relation","to","strength")]
  names(arcs) = c("Parent","-","Child","Weight")
  write.table(arcs,file=file, row.names = FALSE, quote = FALSE,sep=" ")
  nodeTimes = list("names"=names(network$nodes),"times"=apply(as.matrix(unlist(names(network$nodes))), MARGIN=1, FUN=function (x) {as.numeric(strsplit(x,"_t")[[1]][2])}))
  write.table(nodeTimes,paste0(file,".timetable"), row.names = FALSE, quote = FALSE, sep=" ")
}

write.sif(paste0("processed-data/",cell.type,"/DBN.sif"), network, df)
