---
title: "constructPseudotemporalOrdering.Rmd"
output:
  html_notebook: default
  pdf_document: default
---
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/home/samsl/Pseudotemporal-GRNs")
knitr::opts_chunk$set(echo = TRUE)
```

```{r "Load Libraries", warning=FALSE, include=FALSE, results='hide'}
require(ACTIONet)
require(SingleCellExperiment)
require(Seurat)
#require(monocle3)
require(ComplexHeatmap)
require(factoextra)
require(ggfortify)
require(xlsx)
require(UsingR)
```

### Load SCE and ACTIONet for a specific cell type
Choose from cell types:
  - Astrocytes: "Ast"
  - Excitatory Neurons: "Ex"
  - Inhibitory Neurons: "In"
  - Microglia: "Mic"
  - Oligodendrocytes: "Oli"
  - Oligodendrocyte Progenitor Cells: "OPC"
  
Use HD Microglia:
  - "HD_mic"
```{r "Load Data"}
# Select cell type from list
cell.type = "Mic"
sce.file.name <- paste("/home/samsl/Pseudotemporal-GRNs/processed-data/",cell.type,"/",cell.type,"_sce.rds",sep="")
ACTIONet.file.name <- paste("/home/samsl/Pseudotemporal-GRNs/processed-data/",cell.type,"/",cell.type,"_ACTIONet.rds",sep="")
sce <- readRDS(sce.file.name)
ACTIONet.out <- readRDS(ACTIONet.file.name)
```

# Identify Cell States Correlated with Disease
To produce a pseudotemporal ordering of cell states, we identify cell states which are correlated with disease. We can use `braaksc`, `ceradsc`, or `cogdx` as clinical indicators of disease progression. For now, we will use `ceradsc`. We plot enrichment for clinical phenotype across cell states to identify the maximally enriched cell state.
```{r "Alz Find Cluster Markers"}
# Import Clinical Data from ROSMAP
clinical <- read.table("/home/samsl/Pseudotemporal-GRNs/raw-data/Processed_10x_AD_data/ROSMAP_Clinical_2019-05_v3.csv", header = TRUE, sep=",")

cogdx.ind <- factor(unlist(lapply(colData(sce)$projid, function (x) { clinical[which(clinical$projid == x),"cogdx"] })))
braak.ind <- factor(unlist(lapply(colData(sce)$projid, function (x) { clinical[which(clinical$projid == x),"braaksc"] })))
cerad.ind <- factor(unlist(lapply(colData(sce)$projid, function (x) { clinical[which(clinical$projid == x),"ceradsc"] })))
indicators = list("braak" = braak.ind, "cogdx" = cogdx.ind, "cerad" = cerad.ind)

# Select Indicator to Use
selected.ind = "braak"

# Annotate ACTIONet with each clinical indicator
ACTIONet.out <- add.cell.annotations(ACTIONet.out, indicators[[selected.ind]], annotation.name=selected.ind)

# Visualize  the clinical indicators in the ACTIONet layout
plot.ACTIONet(ACTIONet.out, labels = ACTIONet.out$annotations[[selected.ind]]$Labels)

# Identify disease progression enriched archetypes
annot.out <- annotate.archetypes.using.labels(ACTIONet.out, indicators[[selected.ind]], core = TRUE)
svg(paste("img/",cell.type,"_cell_state_",selected.ind,"_enrichment.svg",sep=""))
Heatmap(t(annot.out$Enrichment), cluster_rows = FALSE, cluster_columns = FALSE, name='Enrichment', column_title='Cell State', column_title_side='bottom')
```

```{r "HD Find Cluster Markers"}
# Import Clinical Data from ROSMAP
ACTIONet.out <- HD.dat$ACTIONet
clinical <- read.xlsx("/home/samsl/Pseudotemporal-GRNs/raw-data/Processed_HD_data/Human_HD_metedata.xlsx", 1, header=TRUE)
clinical$Grade <- as.factor(as.numeric(clinical$Grade))
clinical <- clinical[1:12,]

grade.ind <- factor(unlist(lapply(colData(HD.dat$sce)$batch, function (x) { clinical[which(clinical$Batch.Order == x),"Grade"] })))

# Annotate ACTIONet with each clinical indicator
ACTIONet.out <- add.cell.annotations(ACTIONet.out, HD.dat$sce$grade, annotation.name="grade")

# Visualize  the clinical indicators in the ACTIONet layout
plot.ACTIONet(ACTIONet.out, labels = ACTIONet.out$annotations[["grade"]]$Labels)

# Identify disease progression enriched archetypes
annot.out <- annotate.archetypes.using.labels(ACTIONet.out, HD.dat$sce$grade, core=TRUE)
svg(paste("img/",cell.type,"_cell_state_",selected.ind,"_enrichment.png",sep=""))
Heatmap(t(annot.out$Enrichment), cluster_rows = FALSE, cluster_columns = FALSE, name='Enrichment', column_title='Cell State', column_title_side='bottom')
```

# Biological Pathway Enrichment
We can verify the correlation of this cell state with progression by searching for enrichment with GO pathways.
```{r "Pathway Enrichment"}
data("nanoStringDB_human")
ADmarkers <- nanoStringDB_human$AD
enrichment.out <- t(geneset.enrichment.archetype(ACTIONet.out, ADmarkers))
svg(paste0("img/",cell.type,"_cell_state_disease_enrichment.svg"))
Heatmap(t(enrichment.out), cluster_columns=FALSE, name='Enrichment', column_title = 'Cell State', column_title_side = 'bottom')
```

# Calculate Disease Score (Alz)
This cell state can then be used to assign scores for each cell.
```{r "Alz Disease Score"}
# Indicators vote for most correlated state
MAX.BRAAK.STAGE <- 6
braak.ind.enrich <- annotate.archetypes.using.labels(ACTIONet.out, indicators[["braak"]], core = TRUE)$Enrichment
braak.vote <- which.max(braak.ind.enrich[,MAX.BRAAK.STAGE])

MAX.CERAD.STAGE <- 1
cerad.ind.enrich <- annotate.archetypes.using.labels(ACTIONet.out, indicators[["cerad"]], core = TRUE)$Enrichment
cerad.vote <- which.max(cerad.ind.enrich[,MAX.CERAD.STAGE])

MAX.COGDX.STAGE <- 5
cogdx.ind.enrich <- annotate.archetypes.using.labels(ACTIONet.out, indicators[["cogdx"]], core = TRUE)$Enrichment
cogdx.vote <- which.max(cogdx.ind.enrich[,MAX.COGDX.STAGE])

DISEASE.ENRICH.COL <- 1
disease.enrich <- t(geneset.enrichment.archetype(ACTIONet.out, ADmarkers))
disease.vote <- which.max(disease.enrich[,DISEASE.ENRICH.COL])

ACTIVE.MIC.COL <- 2
active.mic.vote <-  which.max(disease.enrich[,ACTIVE.MIC.COL])


listMode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

correlated.state <- listMode(c(braak.vote, cerad.vote, cogdx.vote, disease.vote, active.mic.vote))

disease.score <- ACTIONet.out$unification.out$H.core[correlated.state,]
core.significance <- ACTIONet.out$unification.out$DE.core@assays$data$significance
associated.genes <- rownames(core.significance)[order(core.significance[,correlated.state], decreasing=TRUE)]
svg(paste("img/",cell.type,"_archetype_disease_score_gradient.svg",sep=""))
plot.ACTIONet.gradient(ACTIONet.out,disease.score)
```

# Calculate Disease Score (HD)
This cell state can then be used to assign scores for each cell.
```{r "HD Disease Score"}
# Indicators vote for most correlated state
MAX.GRADE <- 4
grade.enrich <- annotate.archetypes.using.labels(ACTIONet.out, grade.ind, core = TRUE)$Enrichment
grade.vote <- which.max(grade.enrich[,MAX.GRADE])

DISEASE.ENRICH.COL <- 1
disease.enrich <- t(geneset.enrichment.archetype(HD.dat$ACTIONet, HD.markers$genes))
disease.vote <- which.max(disease.enrich[,DISEASE.ENRICH.COL])

ACTIVE.MIC.COL <- 2
active.mic.vote <-  which.max(disease.enrich[,ACTIVE.MIC.COL])


listMode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

correlated.state <- listMode(c(grade.vote, disease.vote, active.mic.vote))
correlated.state <- grade.vote

disease.score <- ACTIONet.out$unification.out$H.core[correlated.state,]
core.significance <- ACTIONet.out$unification.out$DE.core@assays[["significance"]]
associated.genes <- rownames(core.significance)[order(core.significance[,correlated.state], decreasing=TRUE)]
png(paste("img/",cell.type,"_archetype_disease_score_gradient.png",sep=""))
plot.ACTIONet.gradient(ACTIONet.out,disease.score)
```

# View Biological Pathways enriched in the correlated archetype
We can see which pathways are enriched in the correlated state and other states.
```{r}
data("gProfilerDB_human")
gpADmarkers <- gProfilerDB_human$SYMBOL$`GO:BP`
gp.enrichment.out <- geneset.enrichment.archetype(ACTIONet.out, gpADmarkers)
gp.top <- gp.enrichment.out[order(gp.enrichment.out[,correlated.state], decreasing = TRUE),][1:20,]
svg(paste0("img/",cell.type,"_cell_state_pathway_enrichment.svg"))
Heatmap(gp.top, cluster_columns=FALSE, name='Enrichment', column_title='Cell State', column_title_side='bottom')
```

# Labels from archetypes and progression
```{r "Label cells based on enrichment for disease progression" }
ACTIONet.out <- annotate.cells.from.archetype.enrichment(ACTIONet.out,t(annot.out$Enrichment),core = TRUE,annotation.name = paste0(cell.type,".enrichment.labels"))

png(paste0("/home/samsl/Pseudotemporal-GRNs/img/",cell.type,"_cell_labels_",selected.ind,"_enrichment_labels.png"))
plot.ACTIONet(ACTIONet.out, labels=paste0(cell.type,".enrichment.labels"))
```

# Label cells based on disease score
```{r}
png(paste0("img/",cell.type,"_archetype_cell_labels.png"))
plot.ACTIONet(ACTIONet.out, ACTIONet.out$unification.out$assignments.core)
plot.ACTIONet.3D(ACTIONet.out, ACTIONet.out$unification.out$assignments.core)
```

```{r}
nbins = 4

disc = discretize(as.data.frame(disease.score),"interval",breaks=nbins,ordered=TRUE)
disc$raw.score = disease.score

ggplot(disc, aes(x=raw.score)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white",binwidth = 1/nbins, boundary = 0, closed = "left")+
 geom_density(alpha=.2, fill="#FF6666") +
geom_rug(aes(y = 0), position = position_jitter(height = 0)) +
  coord_cartesian(xlim=c(0, 1)) +
  ggtitle("Density Histogram of Disease Score") +xlab("Disease Score")
ggsave(paste0("img/",cell.type,"_density_hist.png"))
```

```{r}
#Cluster cells by disease score
M.genes = dim(sce)[1]
top.genes = associated.genes[1:M.genes]
top.sce <- sce[rownames(sce) %in% top.genes]
df <- as.data.frame(as.matrix(t(counts(top.sce))))

K.centers = nbins
km <- kmeans(df, K.centers)
pc <- prcomp(t(df))
plot(pc$rotation[,1],disease.score, col=km$cluster, xlab="PC1")
#png(paste0("img/",cell.type,"_disease_pc1.png"))
plot(pc$rotation[,1],disease.score, col=disc$disease.score, xlab="PC1")

clusts <- as.data.frame(list("CellNames" = colnames(sce),
                             "KMeans" = km$cluster,
                             "DiseaseScore" = disease.score,
                             "DiseaseRange" = disc$disease.score,
                             "ArchetypeCluster" = ACTIONet.out$unification.out$assignments.core
                             ))

saveRDS(df,paste0("processed-data/",cell.type,"/top_gene_counts.rds"))
saveRDS(clusts,paste0("processed-data/",cell.type,"/clusters.rds"))
```